A generalized definition of multidimensional item response theory parameters

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects repeated uses of "Therefore"

---
title: "A generalized definition of multidimensional item response theory parameters"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
abstract: |
  Modern measurement theory demands that Multidimensional Item Response Theory
  accurately represents covariate spaces with correlated dimensions and/or
  non-standard variances. As it is also used in applications in the
  non-cognitive domain, they also need to model negatively-keyeed items
  accurately and interpretably. Henceforth we generalize the
  multidimensional discrimination [@reckase_discriminating_1991] and
  difficulty [@reckase_difficulty_1985] parameters to account for those two
  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
  Model, represented in a latent space with an arbitrary basis, and
  then defining such basis to be a concrete geometrical representation of the
  phenomena we intend to measure. This results in three different versions of
  the multidimensional parameters: The original one, based on the item
  parameters solely, one that incorporates the covariance structure of the
  latent space, and a third one that uses the correlation structure instead.
  Following @ackerman_creating_1994, we also provide a geometrical
  representation based on these parameters, and provide applied examples to
  illustrate the pertinence of these generalized parameters to represent the
  measurement properties of the items accurately. We recommend using the
  version that uses the latent covariance structure for describing the
  properties of the parameters, and the one based on the correlation structure
  for graphical representation. Finally, we discuss the implications of this
  generalization for other multidimensional IRT models, and the parallels of our
  results in common factor model theory.
keywords: |
  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
  item discrimination, item difficulty, item vector representation
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

@@

**Keywords:** Multidimensional IRT, 2-Parameter Logistic model,
multidimensional parameters, item discrimination, item difficulty,
item vector representation

@@

The item location $`r MIL_PARAM`$ is defined as the distance and direction
from the origin to the point of maximum slope [@reckase_difficulty_1985].
By analogy with the unidimensional case, $`r MDISC_SYM`$ is defined as
$4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,
the choice we make of this matrix will lead us to different definitions of
the parameters.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
Therefore, for graphical representation purposes,
the correlation-based version will be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we will need to compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Therefore, we will need to define
a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, we have discussed that this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and (2) to any case of unchanging monotony.
The latter may seem superfluous, given that several examples in the literature
already make use of this generalization
[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
However, this generalization had never been formalized, up to now.
Therefore, we highlight here the interest of showing how
the item psychometric properties are paralleled by their geometrical properties,
and how this holds for any case of unchanging monotony
of the item response function.
Regarding the first generalization,
it is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotony).
However, they provided no formal proof for this result either.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

@@

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is important in applications where
potentially different bases are involved.
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad-hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
This implies that the axes defined by the test space basis,
and not the latent space basis,
are a geometrical representation of the covariance structure.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
We hope our results help shed a clearer light on
this often misunderstood topic.

It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although proposals such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying multidimensional measurement theory, its implications,
and its interpretation.

>>>>

---
title: "A generalized definition of multidimensional item response theory parameters"
author: "Daniel Morillo"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

@@

@@

The item location $`r MIL_PARAM`$ is defined as the distance and direction
from the origin to the point of maximum slope [@reckase_difficulty_1985].
By analogy with the unidimensional case, $`r MDISC_SYM`$ is defined as
$4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
Hence we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,
the choice we make of this matrix will lead us to different definitions of
the parameters.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, we have discussed that this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and (2) to any case of unchanging monotony.
It is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotony).
However, they provided no formal proof for this result.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

One may consider that the formal derivations presented here are overly
complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Moreover, strictly following @reckase_difficulty_1985's method in the original,
oblique space has also provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although models such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying measurement theory, its implications and its interpretation.
